---
title: Benchmarks
description: Yamark performance against other YAML and Markdown formatters.
---

```{r}
#| label: benchmark-data
#| include: false

source("_benchmark-data.R")
```

`r bigfile_headline()` `r directory_headline()`

There is one comparison per input kind, and each lists every tool whose own
CLI formats that input natively, used simply: no formatting options, no
shims, no adapters. The tool roster therefore differs by input kind.

::: {.panel-tabset}

## Markdown

One generated 4 MB Markdown document (`big.md`): prose paragraphs with
links - some longer than the line width - and nested lists with mixed
markers. Each tool's CLI formats the file in place; time includes process
startup. Time is `r measurement_inline(big_target_rows(markdown_target))`,
on a fresh copy of the file each run. Memory is median peak RSS.

```{r}
#| echo: false
#| results: asis
write_table(
  big_table(markdown_target),
  align = c("l", "r", "r", "r")
)
```

## YAML

One generated 4 MB YAML file (`big.yaml`): block maps and sequences,
block scalars, and comments. Same procedure as the Markdown comparison.

```{r}
#| echo: false
#| results: asis
write_table(
  big_table(yaml_target),
  align = c("l", "r", "r", "r")
)
```

## Markdown + front matter

The same 4 MB document shape (`big-with-frontmatter.md`) with a
`r fmt_kb(big_requested_frontmatter_yaml)` deliberately unformatted YAML
front matter block - about a third of the document's lines are YAML. The
Front matter column reports what each tool did with that block:
**formatted** (rewrote it), **untouched** (passed it through), or **not
preserved** (broke the delimiters).

```{r}
#| echo: false
#| results: asis
write_table(
  big_table(frontmatter_target, front_matter = TRUE),
  align = c("l", "r", "r", "c", "r")
)
```

The harness derives that column from the output bytes: trailing-whitespace
trimming does not count as formatting. `dprint`'s Markdown plugin passes
front matter through unformatted, and `mdformat` (installed without its
front-matter plugin) reads the opening `---` as a thematic break and
corrupts the block.

## Directory

`r flow_directory_rows$files[[1]]` generated YAML service-configuration
files of about 100 KB each
(`r fmt_mb_round(flow_directory_rows$corpus_bytes[[1]])` in total). Each
tool is passed the directory root once and discovers the files itself;
every run is verified to have reformatted all
`r flow_directory_rows$files[[1]]` files. Time is
`r measurement_inline()`, on a fresh copy of the corpus each run.

```{r}
#| echo: false
#| results: asis

directory_table <- data.frame(
  Formatter = flow_directory_rows$formatter,
  Time = fmt_sec(flow_directory_rows$median_seconds),
  `User CPU` = fmt_sec(flow_directory_rows$median_user_seconds),
  Throughput = fmt_mbps(flow_directory_rows$mb_per_second),
  `vs yamark` = vapply(
    flow_directory_rows$median_seconds,
    slower_than,
    character(1),
    base = directory_yamark_seconds
  ),
  check.names = FALSE
)
write_table(directory_table, align = c("l", "r", "r", "r", "r"))
```

Yamark formats the directory in parallel (as do `deno-fmt` and `dprint`),
so wall time can beat user CPU time; the User CPU column is the
single-core comparison, and yamark is fastest on that column too.
Throughput is input MB/s.

:::

## How to read these results

Every number comes from the same harness (`tools/bench/big.py` and
`tools/bench/run.py` in the repository), running each tool the way a user
would: its own CLI, default configuration, no formatting options, against
deterministic generated corpora. The comparison set is the other formatters
in the space - `r compared_tools_inline()`.

Two harness details keep that comparison clean without changing how any
tool is invoked:

- `panache` and `dprint` cache format results. The harness redirects each
  tool's cache into the benchmark work directory and clears it between
  repetitions, so every timed run formats from scratch rather than
  replaying a cached result.
- `dprint` has no built-in plugins, so it runs with a config file that
  names its first-party plugin for the input kind (Markdown or YAML) and
  sets nothing else.

Lint fixers with an autofix mode (`pymarkdown`, `markdownlint-cli2`) are
not formatters, so they are not part of the comparison. The harness can
still run them via `--tools`.

Measured on a MacBook Pro (`r host_inline()`). `r tool_versions_block()`

The tables render the latest checked-in artifacts -
`r artifact_link(big_ok_rows)` for the single-file comparisons and
`r artifact_link(flow_directory_rows)` for the directory comparison - which
record the full per-run timings, output hashes, git commit, and host
details. A table renders only if every tool in its roster completed the
benchmark; degraded runs are never shown as smaller tables.

## Reproducing

```sh
tools/bench/big.py
tools/bench/run.py --invocation directory --files 500 --items 540 \
  --reps 3 --warmups 1 \
  --tools yamark,yamlfmt,prettier,yamlfix,dprint-yaml,deno-fmt
```

The corpora are generated deterministically, so the same commands reproduce
the same inputs anywhere; each script writes a JSON artifact under
`docs/benchmarks/`.
