---
title: Command line
description: Modes, output, exit status, and generated command-line help.
execute:
  echo: false
  warning: false
  message: false
  comment: ""
---

```{r}
#| label: setup-cli-help
#| include: false
source("_yamark-build.R")

yamark_help <- function(...) {
  args <- c(..., "--help")

  output <- system2(
    yamark_bin,
    args,
    stdout = TRUE,
    stderr = TRUE,
    env = c("NO_COLOR=", "CLICOLOR_FORCE=1")
  )
  status <- attr(output, "status")

  if (!is.null(status)) {
    stop(
      "`", yamark_bin, " ", paste(args, collapse = " "), "` failed with status ", status,
      call. = FALSE
    )
  }

  output <- sub("[ \t]+$", "", output, perl = TRUE)
  output <- htmltools::htmlEscape(output)
  output <- fansi::sgr_to_html(output, term.cap = "256")
  knitr::asis_output(paste0(
    "<pre class=\"yamark-cli-help\"><code>",
    paste(output, collapse = "\n"),
    "</code></pre>\n"
  ))
}
```

Use [Usage](usage.qmd) for common commands and
[Formatting settings](reference-options.qmd) to find a control by the output it
changes. This command-line interface (CLI) reference includes help screens
generated from the current Yamark binary.

## Modes, output, and status

| Mode | Files | stdout | stderr |
| --- | --- | --- | --- |
| `yamark format PATHS` | Writes changed files. | Summary. | Failures and requested diagnostics. |
| `yamark format --check PATHS` | Does not write. | - | Summary, failures, and requested diagnostics. |
| `yamark format --diff PATHS` | Does not write. | Unified diffs. | Summary, failures, and requested diagnostics. |
| `yamark format --stdin-file-path PATH` | Does not read or write `PATH`; uses it to select file-aware behavior for stdin. | Formatted content only. | Failures and requested diagnostics. |
| `yamark render --stdin-file-path PATH` | Does not read or write `PATH`; renders stdin using its native or JSON-family file type. | Read-only formatted content only. | Failures. |

`--check` and `--diff` exit `1` when any selected file would change. Every mode
exits `1` when formatting fails and `0` on success otherwise. Invalid command
syntax exits `2`.

When `PATHS` is omitted from `yamark format`, Yamark uses the current directory
(`.`). Format path mode counts unsupported extensions as skipped without
failing. An unsupported `--stdin-file-path` is an error because stdin mode
requires a supported file type.

Format stdin mode rejects additional paths and cannot be combined with
`--check` or `--diff`. `yamark render` always reads stdin, requires
`--stdin-file-path`, and never writes a file.

## Generated help

### `yamark`

```{r}
#| label: help-yamark
#| results: asis
yamark_help()
```

### `yamark format`

```{r}
#| label: help-yamark-format
#| results: asis
yamark_help("format")
```

### `yamark render`

```{r}
#| label: help-yamark-render
#| results: asis
yamark_help("render")
```

### `yamark git-filter`

```{r}
#| label: help-yamark-git-filter
#| results: asis
yamark_help("git-filter")
```
