---
# fmt: skip file
title: Examples
description: >-
  Generated before-and-after examples for supported files and directives.
---

<!-- fmt: skip file -->

```{r}
#| label: setup-showcase-examples
#| include: false
source("_yamark-build.R")

showcase_before_after <- function(before, lang, stdin_file_path, args = character()) {
  before <- trim_one_trailing_newline(before)
  after <- yamark_format(before, stdin_file_path, args = args)

  writeLines(c(
    ":::: {.showcase-before-after}",
    "**Before**",
    "",
    markdown_fence(before, lang),
    "",
    "**After**",
    "",
    markdown_fence(after, lang),
    "::::"
  ))
}
```

These examples run the public Yamark command during the site build. Each one
shows a focused input and the output Yamark writes.

## YAML and Markdown documents

### Markdown with YAML front matter

Agent skills and prompt files can put structured metadata above free-form
Markdown. Yamark formats both in one pass.

```{r}
#| label: showcase-markdown-frontmatter
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(---
name: review-pr
description: Review a pull request and recommend changes inline with the project's style guide, focusing on correctness, readability, and tests.
tags: [review,pull-request,code]
---

#   Review ##

Read the diff and flag anything that violates the style guide, has obvious correctness issues, or looks untested. Prefer specific suggestions over vague concerns.
)---",
  lang = "markdown",
  stdin_file_path = "example.md"
)
```

### Markdown-valued YAML scalars

A `prompts.yaml` or `agents.yaml` file can store Markdown in selected scalar
values. Tag a value `!markdown` or `!md` (or add `# fmt: markdown`) and Yamark
wraps it as Markdown, choosing folded or literal block style from the content.

```{r}
#| label: showcase-markdown-yaml-scalars
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(agents:
  reviewer:
    instructions: !markdown "Focus on correctness and tests. Flag any change that lacks a regression test. Prefer concrete suggestions: name the function, name the case."
  summarizer:
    instructions: !markdown |
      You write release notes.

      - One bullet per user-visible change.
      - No internal refactors.
)---",
  lang = "yaml",
  stdin_file_path = "prompts.yaml"
)
```

The prose-only value is rewrapped as Markdown. The list keeps a literal
block (`|`) because folding would destroy the line breaks Markdown depends on.

### YAML scalar presentation

Yamark changes scalar spelling only when the parsed YAML value and tag
stay equivalent. It can simplify safe quoted strings, keep strings that
would be misread as booleans, convert hard newlines to literal blocks,
and fold long prose strings.

```{r}
#| label: showcase-yaml-scalars
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(title: "hello"
boolish: "true"
body: "alpha\nbeta\n"
summary: "This package formats YAML front matter and configuration files while preserving semantics, comments, and repository-friendly diffs."
)---",
  lang = "yaml",
  stdin_file_path = "scalars.yaml"
)
```

## Markdown in source files

### Embedded Markdown in Python

When a prompt lives next to the code that uses it, Yamark can format the prose
without changing the surrounding Python. Add `# fmt: markdown` before the
string:

```{r}
#| label: showcase-python-markdown-string
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(# fmt: markdown
REVIEW_PROMPT = """
# Review

Read the diff and flag anything that violates the style guide, has obvious correctness issues, or looks untested.

-   Prefer specific suggestions over vague concerns.
-   Name the function and the case.
"""
)---",
  lang = "python",
  stdin_file_path = "prompt.py"
)
```

Python source outside the marked string is left untouched. Run `ruff format` for
that.

### Embedded Markdown in source comments

The same directive can target a contiguous source comment block. This is useful
for generated help text or documentation snippets where a string literal is not
the right host.

```{r}
#| label: showcase-python-markdown-comments
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(# fmt: markdown
# #   Filters
#
# Apply one or more filters. Put the most specific filter first.
#
# -   Each filter is a single expression.
)---",
  lang = "python",
  stdin_file_path = "comments.py"
)
```

### Embedded Markdown in R

The same pattern works for R raw strings. It can format package vignettes,
documentation fragments, and Shiny help text kept next to the code that renders
it.

```{r}
#| label: showcase-r-markdown-string
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(# fmt: markdown
help_text <- r"(
# Filters

Apply one or more filters. Filters are evaluated top-to-bottom; the first match wins, so put the most specific filter first.

-   Each filter is a single expression.
-   Multiple filters combine with AND.
)"
)---",
  lang = "r",
  stdin_file_path = "help.R"
)
```

## YAML layout

### Aligned flow-mapping tables

A sequence of similar flow mappings can be easier to scan as a table. Mark it
with `# fmt: table` and Yamark aligns columns by key:

```{r}
#| label: showcase-flow-table
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(# fmt: table
- {name: alpha,type: string,default: [one,two],description: "Short"}
- {name: beta,type: int,default: 0,description: "Number"}
)---",
  lang = "yaml",
  stdin_file_path = "table.yaml"
)
```

### Compact collections

Short, simple block collections can read better as a single flow line.
Enable `compact = true` in `yamark.toml` (or pass `--compact`) and
Yamark collapses eligible block mappings and sequences that fit the
structural width:

```{r}
#| label: showcase-compact-collections
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(tags:
  - yaml
  - markdown
  - docs
package:
  name: yamark
  language: rust
)---",
  lang = "yaml",
  stdin_file_path = "compact.yaml",
  args = "--compact"
)
```

Collections with comments, aliases, tags, anchors, multiline scalars, or
block scalars stay in block style.

### Collapse to flow by typing a bracket

Drop an unmatched `[` or `{` where a block sequence or mapping value starts and
Yamark reads it as a layout hint: collapse this collection onto one line. The
file doesn't parse as you typed it, but the intent is obvious - Yamark removes
the opener and emits flow style.

```{r}
#| label: showcase-collapse-inline-bracket
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(tags: [
  - yaml
  - markdown
  - docs
)---",
  lang = "yaml",
  stdin_file_path = "repair.yaml"
)
```

The opener may also be detached onto the next line after an empty mapping value
header:

```{r}
#| label: showcase-collapse-detached-bracket
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(tags:
[
  - yaml
  - markdown
  - docs
)---",
  lang = "yaml",
  stdin_file_path = "repair-detached.yaml"
)
```

Adjacent forms such as `tags:[` and `tags:{` are not repaired as layout hints.

A newline inside an existing flow collection requests the reverse change: expand
the collection to block style.

```{r}
#| label: showcase-expand-flow-newline
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(- {a: b,
   c: d}
)---",
  lang = "yaml",
  stdin_file_path = "expand.yaml"
)
```

See [Reference -> Layout repair](reference.qmd#layout-repair) for the
acceptance rules.

## Fenced and nested content

### Recursive Markdown code fences

YAML fences are formatted as YAML. Markdown fences are formatted
recursively, so nested YAML inside nested Markdown is formatted too.

``````{r}
#| label: showcase-recursive-markdown-fences
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(```yaml
# fmt: table
- {name: a,type: int,default: 0}
- {name: long_name,type: string,default: ""}
```

````markdown
#   Inner

```yaml
items: [a,b]
```
````
)---",
  lang = "markdown",
  stdin_file_path = "recursive.md"
)
``````

### Prettier-backed web and data fences

When `prettier` is on `PATH`, JSON, JSONC, JSON5, GraphQL, CSS, SCSS, Less,
PostCSS, HTML, JavaScript, JSX, TypeScript, and TSX fenced blocks are handed to
it. Python and R fences go to Ruff and Air the same way:

````{r}
#| label: showcase-prettier-fences
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(```json
{"name":"yamark","tags":["yaml","markdown","formatter"],"engines":{"node":">=18"}}
```

```ts
function format(input:string):string{return input.trim()}
```
)---",
  lang = "markdown",
  stdin_file_path = "web.md"
)
````

A missing `prettier` leaves the fence unchanged. Override or disable a
language with `[embedded.<language>]` in `yamark.toml`, or skip them all with
`--skip-embedded-formatters`.

### Embedded source in YAML literal blocks

Configure an embedded formatter for a language in `yamark.toml`
(`[embedded.r]`, `[embedded.python]`, custom), then mark a literal
block scalar with `# fmt: r` or `# fmt: python` to format the embedded
code through the external formatter:

```{r}
#| label: showcase-yaml-python-scalar
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(# fmt: python
preflight: |
  def check(items):return all(x>0 for x in items)
)---",
  lang = "yaml",
  stdin_file_path = "embedded-python.yaml"
)
```

Yamark sends the scalar to the configured formatter and re-indents the result in
the YAML block.

### Markdown links, footnotes, and tables

Long inline links are split at Markdown syntax boundaries. Footnote
definitions wrap under the marker unless the document or CLI asks to
preserve them.

```{r}
#| label: showcase-markdown-links-footnotes
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(Please be mindful of our [code of conduct](https://github.com/quarto-dev/quarto-cli/blob/main/.github/CODE_OF_CONDUCT.md) as you interact with other community members.

Body text with a footnote.[^long]

[^long]: This footnote explains that comments were removed because people used comments to hold parsing directives and enough extra words to wrap.
)---",
  lang = "markdown",
  stdin_file_path = "links.md"
)
```

Simple GFM pipe tables are aligned by display width. Git clean/smudge filters
always use compact pipe-table output.

```{r}
#| label: showcase-markdown-table
#| echo: false
#| results: asis
showcase_before_after(
  before = r"---(| package | label |
|---|---|
| dplyr | tidy |
| tidyr | pivoting |
)---",
  lang = "markdown",
  stdin_file_path = "table.md"
)
```

---

See [Reference](reference.qmd) for the full directive list, options,
and safety model.
