---
title: Yamark
description: An extremely fast formatter for YAML and Markdown, written in Rust.
toc: false
---

```{r}
#| label: setup-homepage-examples
#| include: false
source("_yamark-build.R")

get_chunk <- function(label) {
  paste(knitr::knit_code$get(label), collapse = "\n")
}

hero_lines <- function(x) {
  lines <- strsplit(x, "\n", fixed = TRUE)[[1]]
  lines <- lines[grepl("^tags:|^#[[:space:]]", lines)]
  stopifnot(length(lines) == 2)
  lines
}

hero_diff_line <- function(kind, sign, value) {
  label <- if (identical(kind, "remove")) "Before" else "After"
  sprintf(
    paste0(
      '<div class="hero-diff-line %s">',
      '<span class="hero-diff-sign" aria-hidden="true">%s</span>',
      '<span class="visually-hidden">%s: </span>',
      '<code>%s</code>',
      '</div>'
    ),
    kind,
    sign,
    label,
    htmltools::htmlEscape(value)
  )
}

emit_hero_specimen <- function(label) {
  before <- get_chunk(label)
  after <- yamark_format(before, stdin_file_path = "guide.qmd")
  before_lines <- hero_lines(before)
  after_lines <- hero_lines(after)
  stopifnot(!identical(before_lines, after_lines))

  writeLines(c(
    '<div class="hero-diff-body">',
    hero_diff_line("remove", "-", before_lines[[1]]),
    hero_diff_line("add", "+", after_lines[[1]]),
    '<div class="hero-diff-gap" aria-hidden="true"></div>',
    hero_diff_line("remove", "-", before_lines[[2]]),
    hero_diff_line("add", "+", after_lines[[2]]),
    '</div>'
  ))
}

emit_before_after <- function(label, stdin_file_path = "example.md") {
  before <- get_chunk(label)
  after <- yamark_format(before, stdin_file_path = stdin_file_path)

  writeLines(c(
    ":::: {.before-after}",
    "::: {.before-after-pane #demo-before}",
    '**Before** <label class="softwrap-toggle"><input type="checkbox" id="demo-softwrap-toggle"> soft wrap</label>',
    "",
    markdown_fence(before),
    "",
    ":::",
    "",
    "::: {.before-after-pane}",
    "**After**",
    "",
    markdown_fence(after),
    ":::",
    "::::"
  ))
}
```

```{cat}
#| label: hero-example
#| include: false
---
tags: [docs,agents,yaml]
---

# Agent guide ##
```

::: {.hero-shell}
::: {.hero-copy}
::: {.hero-kicker}
![](assets/favicon.svg){.hero-mark}
:::

Markdown and YAML are source files too. Yamark gives them the consistent
formatting we expect from code, whether they appear as standalone files, front
matter, fenced blocks, YAML scalars, or marked strings and comments in
Python&nbsp;and&nbsp;R.

::: {.hero-actions}
[Install](usage.qmd#install){.hero-button .primary}
[See examples](examples.qmd){.hero-button}
[Benchmarks](benchmarks.qmd){.hero-button}
:::

::: {.hero-command}
<span>Run directly from PyPI</span>
`uvx yamark format`
:::
:::

::: {.hero-specimen}
::: {.hero-specimen-header}
<span class="hero-specimen-path">guide.qmd</span>
<span class="hero-specimen-kind">YAML + Markdown</span>
:::

```{r}
#| label: render-hero-example
#| echo: false
#| results: asis
emit_hero_specimen("hero-example")
```
:::
:::

::: {.workflow-strip}
::: {.workflow-item}
**Format**

Rewrite supported files in place.
:::

::: {.workflow-item}
**Check**

Use non-mutating CI and diff modes.
:::

::: {.workflow-item}
**Embed**

Format marked Markdown in YAML, Python, or R.
:::

::: {.workflow-item}
**Recurse**

Format nested YAML and Markdown inside fenced blocks.
:::
:::

## A quick example

Here is a Markdown file with YAML front matter, a long paragraph, and a nested
list. Yamark formats the YAML and Markdown in one pass. The first pane shows the
input; the second shows what `yamark format` writes back.

```{cat}
#| label: example-text
#| include: false
---
title: Why YAML + Markdown?
description: Structured fields for software alongside prose, instructions, examples, and code for readers.
tags: [documentation,configuration,formatting]
---

#   Why YAML + Markdown? ##

YAML front matter and a Markdown body let one file serve readers and software. The front matter carries fields a program can inspect; the body carries prose that people can edit and review in a diff.

This shape is useful for:

- Repository documents that need to render for readers and remain easy for tools to inspect.
  - The body can carry examples and nested Markdown structures.
- Prompts and agent instructions where metadata sits next to free-form text.
- Text that moves between documentation, configuration, and tool input without a separate source format.
```

```{r}
#| label: render-example-text
#| echo: false
#| results: asis
emit_before_after("example-text")
```

Toggle soft wrap on the Before pane to inspect the input's physical lines.
Yamark wraps and indents each region according to its own grammar.

```{=html}
<script>
(function () {
  var toggle = document.getElementById('demo-softwrap-toggle');
  var pane = document.getElementById('demo-before');
  if (!toggle || !pane) return;
  var apply = function () { pane.classList.toggle('soft-wrap', toggle.checked); };
  toggle.addEventListener('change', apply);
  apply();
})();
</script>
```

::: {.feature-grid}
::: {.feature}
### Mixed documents

Format YAML front matter and the Markdown body with one command.
:::

::: {.feature}
### Rewrap after writing

Edit Markdown prose, YAML descriptions, prompt bodies, and front matter without
maintaining line breaks by hand.
:::

::: {.feature}
### Nested content

Format YAML and Markdown recursively, and send supported embedded code to its
own formatter when configured.
:::
:::

## Performance

```{r}
#| label: setup-home-benchmarks
#| include: false
source("_benchmark-data.R")
```

`r bigfile_headline()` `r directory_headline()`

The [Benchmarks](benchmarks.qmd) page has the full tables, methodology, checked-in
results, and reproduction commands.
