---
title: Yamark
body-classes: yamark-home
toc: false
---

```{r}
#| label: setup-homepage-examples
#| include: false
source("_yamark-build.R")

get_chunk <- function(label) {
  paste(knitr::knit_code$get(label), collapse = "\n")
}

emit_before_after <- function(label, stdin_file_path = "example.md") {
  before <- get_chunk(label)
  after <- yamark_format(before, stdin_file_path = stdin_file_path)

  writeLines(c(
    ":::: {.before-after}",
    "::: {.before-after-pane #demo-before}",
    '**Before** <label class="softwrap-toggle"><input type="checkbox" id="demo-softwrap-toggle"> soft wrap</label>',
    "",
    markdown_fence(before),
    "",
    ":::",
    "",
    "::: {.before-after-pane}",
    "**After**",
    "",
    markdown_fence(after),
    ":::",
    "::::"
  ))
}
```

::: {.hero-shell}
::: {.hero-copy}
::: {.hero-kicker}
![](assets/favicon.svg){.hero-mark}
:::

<h1 class="hero-thesis">A fast formatter for YAML and Markdown.</h1>

Yamark formats whole files and embedded content with the consistency we expect
from code, keeping source readable and changes easy to review.

::: {.hero-actions}
[Install](usage.qmd#install){.hero-button .primary}
[See examples](examples.qmd){.hero-button}
:::

::: {.hero-command}
<span>Run directly from PyPI</span>
`uvx yamark format`
:::
:::
:::

::: {.workflow-strip}
::: {.workflow-item}
**Format**

Rewrite supported files in place.
:::

::: {.workflow-item}
**Check**

Use non-mutating CI and diff modes.
:::

::: {.workflow-item}
**Embed**

Format marked Markdown in YAML, Python, or R.
:::

::: {.workflow-item}
**Recurse**

Format nested YAML and Markdown inside fenced blocks.
:::
:::

## A quick example

A short Markdown file with YAML front matter - a native document format for LLM
work. It combines software-readable structure with natural language for prompts,
skills, reference documents, and generated output. The Before pane shows the
unformatted source; the After pane shows what `yamark format` writes back.

```{cat}
#| label: example-text
#| include: false
---
title: Why YAML + Markdown?
description: The source languages of LLM work - structured fields for tools alongside instructions, examples, and context for models.
tags: [llm,authoring,formats]
---

#   Why YAML + Markdown? ##

When LLMs become part of a software system, YAML and Markdown start to feel like programming languages. YAML defines named fields tools can act on; Markdown carries the instructions, examples, and context that shape model behavior. Together they create a shared source file that stays readable to people and models, exposes explicit structure to software, produces clear diffs, and renders directly.

Where the combination shines:

- Agent skills and prompt files keep names, versions, and tool metadata beside the instructions they describe.
  - The Markdown body can carry code samples and nested structures with their own syntax and indentation.
- Reference documents stay readable to authors and models without a separate source format or build step.
- Tool inputs and model outputs survive chat, Git, and documentation pipelines as recognizable text.
- One source can render as HTML or PDF and travel unchanged as agent context.
```

```{r}
#| label: render-example-text
#| echo: false
#| results: asis
emit_before_after("example-text")
```

Toggle soft wrap on the Before pane to inspect the input's physical lines.
Yamark wraps and indents each region according to its own grammar.

```{=html}
<script>
(function () {
  var toggle = document.getElementById('demo-softwrap-toggle');
  var pane = document.getElementById('demo-before');
  if (!toggle || !pane) return;
  var apply = function () { pane.classList.toggle('soft-wrap', toggle.checked); };
  toggle.addEventListener('change', apply);
  apply();
})();
</script>
```

## Performance

```{r}
#| label: setup-home-benchmarks
#| include: false
source("_benchmark-data.R")
```

`r benchmark_headline()`

```{r}
#| label: homepage-benchmark-field
#| echo: false
#| results: asis

write_benchmark_chart(
  id = "homepage-benchmark-field",
  title = "Elapsed time across formatter CLIs",
  subtitle = paste(
    "Every formatter in the checked-in comparison.",
    "Median wall time; lower is better.",
    "All four panels share a logarithmic seconds scale."
  ),
  rows = benchmark_full_field_rows()
)
```

The [Benchmarks](benchmarks.qmd) page includes detailed tables, methodology,
checked-in results, and reproduction commands.
