---
title: Yamark
description: A formatter for YAML and Markdown.
toc: false
---

::: {.hero-shell}
::: {.hero-copy}
::: {.hero-kicker}
![](assets/favicon.svg){.hero-mark}
[Beta]{.status-chip}
:::

Yamark is a command-line formatter for YAML and Markdown. It formats YAML front
matter and supported fenced blocks alongside Markdown prose. It can also format
explicitly marked Markdown in YAML, Python, and R files. Optional formatters
such as Ruff, Air, and Prettier handle their matching fenced code blocks when
installed.

::: {.hero-actions}
[Install](usage.qmd#install){.hero-button .primary}
[See examples](examples.qmd){.hero-button}
[Benchmarks](benchmarks.qmd){.hero-button}
:::
:::

::: {.terminal-window}
::: {.terminal-chrome}
<span>uvx yamark format</span>
:::

```sh
$ uvx yamark format config.yaml docs/
format config.yaml
format docs/index.md
format docs/reference.qmd
```
:::
:::

::: {.workflow-strip}
::: {.workflow-item}
**Format**

Rewrite supported files in place.
:::

::: {.workflow-item}
**Check**

Use non-mutating CI and diff modes.
:::

::: {.workflow-item}
**Embed**

Format marked Markdown in YAML, Python, or R.
:::

::: {.workflow-item}
**Fences**

Format supported fenced code blocks.
:::
:::

```{r}
#| label: setup-before-after
#| include: false
get_chunk <- function(label) {
  paste(knitr::knit_code$get(label), collapse = "\n")
}

markdown_fence <- function(x, lang = "markdown") {
  runs <- regmatches(x, gregexpr("`+", x, perl = TRUE))[[1]]
  fence <- strrep("`", max(3, max(nchar(runs), 0) + 1))

  paste0(fence, lang, "\n", x, "\n", fence)
}

run_yamark <- function(md, stdin_file_path = "example.md") {
  input <- strsplit(md, "\n", fixed = TRUE)[[1]]

  out <- system2(
    "yamark",
    args = c("format", "--wrap", "72", "--stdin-file-path", stdin_file_path),
    input = input,
    stdout = TRUE,
    stderr = FALSE
  )

  status <- attr(out, "status")
  if (!is.null(status) && !identical(status, 0L)) {
    stop("yamark exited with status ", status, call. = FALSE)
  }

  paste(out, collapse = "\n")
}

emit_before_after <- function(label, stdin_file_path = "example.md") {
  before <- get_chunk(label)
  after <- run_yamark(before, stdin_file_path = stdin_file_path)

  writeLines(c(
    ":::: {.before-after}",
    "::: {.before-after-pane #demo-before}",
    '**Before** <label class="softwrap-toggle"><input type="checkbox" id="demo-softwrap-toggle"> soft wrap</label>',
    "",
    markdown_fence(before),
    "",
    ":::",
    "",
    "::: {.before-after-pane}",
    "**After**",
    "",
    markdown_fence(after),
    ":::",
    "::::"
  ))
}
```

## A quick example

A Markdown file with YAML front matter. The first pane is the input; the second
is what `yamark format` writes back.

```{cat}
#| label: example-text
#| include: false
---
title: Project notes
description: Notes for the next Yamark release.
tags: [yaml,markdown,cli]
---

#   Project notes ##

Yamark formats YAML front matter and the Markdown body in one pass. It wraps long paragraphs, normalizes list indentation, and formats supported fenced code blocks.

-  Run `yamark format` to write changes.
- Run `yamark format --check` in CI.
```

```{r}
#| label: render-example-text
#| echo: false
#| results: asis
emit_before_after("example-text")
```

Toggle soft wrap on the Before pane to inspect the input's physical lines.
Yamark wraps YAML and Markdown separately, then writes one combined document.

```{=html}
<script>
(function () {
  var toggle = document.getElementById('demo-softwrap-toggle');
  var pane = document.getElementById('demo-before');
  if (!toggle || !pane) return;
  var apply = function () { pane.classList.toggle('soft-wrap', toggle.checked); };
  toggle.addEventListener('change', apply);
  apply();
})();
</script>
```

::: {.feature-grid}
::: {.feature}
### Mixed documents

Format YAML front matter and the Markdown body with one command.
:::

::: {.feature}
### Rewrap after writing

Edit Markdown prose, YAML descriptions, and front matter without maintaining
line breaks by hand.
:::

::: {.feature}
### Optional verification

Use `--verify` to reparse changed YAML regions and reject invalid or
value-changing output before writing.
:::
:::

## Performance

```{r}
#| label: setup-home-benchmarks
#| include: false
source("_benchmark-data.R")
```

`r bigfile_headline()` `r directory_headline()`

The [Benchmarks](benchmarks.qmd) page has the full tables, methodology, checked-in
results, and reproduction commands.
